\bf ``A p-value of $<$ 0.05 was considered statistically significant'': a content analysis of published statistical methods sections

### 2021-08-03 manuscript/asa_template/asa_template_almostdone.tex
updated results

An ideal statistical analysis will use appropriate methods to draw
insights from the data and inform the research questions. Unfortunately
many current statistical analyses are far from ideal, with 
researchers often using the wrong methods, misinterpreting the results, or
failing to adequately check their assumptions \citep{2008, Leek2017}.
Some researchers take a ``mechanistic'' approach to statistics, copying
the few methods they know regardless of their appropriateness, and then
going through the motions of the analysis \citep{Stark2018}. Accepting methodological illiteracy in favour of 
research quantity and novelty is at odds
with principles of scientific inquiry, yet continues to pervade published scientific research \citep{VanCalster2021}. 
This paradox has been exemplified during the COVID-19 pandemic, which has led to unprecendented levels of published research of largely 
poor quality due to biases in conduct and reporting \citep{Glasziou2020,Wynants2020}.

@@

We used two openly available data sources to find statistical methods
sections: study
protocols registered on the Australian and New Zealand Clinical Trials
Registry (ANZCTR) and research articles published in \emph{PLOS ONE}. Data sources were chosen as examples of common
research outputs that include descriptions of statistical methods that
were planned or used for analysing study outcomes.

@@

All \emph{PLOS ONE} articles are freely accessible via the PLOS
Application Programming Interface (API). This enabled us to conduct
searches of full-text articles and analyse data on articles' text
content and general attributes such as publication date and field(s) of
research. We applied a two-step
approach to identify statistical methods sections:

@@

Evidence of boilerplate text was assessed at the section and sentences levels using the Jaccard similarity index. 
We chose the Jaccard index as an easy to interpret measure, which summarises the similarity between two pieces of text by the
number of words common to each text (intersection), divided by the
number of words that appeared in either text (union). For both section-level and sentence-level analyses, text was tokenised at the word level. 
Instances of boilerplate text were defined by a Jaccard index of 0.75 or higher and a difference in word count of plus or minus three words.

At the document level, we applied minhash and locality-sensitive hashing using the R package `textreuse' \citep{textreuseR}. The combination of these approaches 
allowed us to reduce the total number of pairwise comparisons within a topic by ignoring unlikely matches. For all topics, locality-sensitive hashing was applied using the same random seed, 
with 1,000 randomly selected minhashes broken into 100 equal-sized bands.

A similar approach was applied at the sentence level. In this case, we aimed to identify common sentences or phrases within statistical methods sections that were representative of each topic. 
First, keyword searches using dominant words from each topic cloud were completed to identify common variations on sentences involving a keyword. Second, we analysed the reuse of sentences from the top 100 sections assigned to each topic, 
based on estimated coherence scores. For each example sentence, text was compared with all sections assigned to the same topic.

@@

We downloaded 28,008 studies. The numbers of excluded studies are shown
in Figure \ref{fig:consort-diagrams}. Of the 12,700 included studies,
9,523 (75\%) had a statistical methods section. The median length of
sections was 129 words with an inter-quartile range of 71 to 219 words.

{\centering \includegraphics[width=0.49\linewidth]{figures/excluded_anzctr_missing}  \includegraphics[width=0.49\linewidth]{figures/excluded_plosone}

\caption{\label{fig:consort-diagrams}Search results for ANZCTR (left) and PLOS ONE (right).}
\end{figure}

@@

{\centering \includegraphics[width=0.7\linewidth]{figures/anzctrwordclouds}

\caption{Word clouds for ten topics for statistical methods sections published in ANZCTR}\label{fig:anzctr-wordclouds}
\end{figure}

Figure \ref{fig:anzctr-wordclouds} shows word clouds for the ten topics identified by the NMF algorithm. 
Words related to the largest topic (topic 2, n = 1,834) reflected descriptions of sample size calculations, for example, ``sample'',``power'' and ``80'' (i.e. $80\%$ power) . Other topics indicated pilot studies (topic 5, n = 834),
safety/tolerability studies (topic 6, n = 524), descriptive analysis (topic 7, n = 333),
intervention studies (topic 8, n = 1020) and repeated measures ANOVA (topic 10 = 852). A review of word counts by topic 
identified statistical methods sections that were only one word, including ``ANOVA'',
``t-test'', ``SPSS'' and even ``SSPS''.

Common statistical methods for data analysis were t-tests (topic 3, n = 1,277), descriptive statistics (topic 7, n = 333), linear models (topic 9, n = 1,195) and repeated measure ANOVA (topic 10, n = 852).
Instances of boilerplate text were concentrated in Topic 3, which emphasised the use of parametric versus non-parametric hypothesis tests (Table \ref{tab:anzctr-example-boilerplate}). Identified matches across topics 7, 9 and 10 were
relatively low, however, sensitiivty in the Jaccard index to shorter sentence lengths may have been a contributing factor. For example, the section with the highest coherence score in topic 7 simply stated `descriptive statistics', with 19 matches. Further review of 
sections assigned to this topic identified a further 7 records with `descriptive analysis' as the entire methods section (Jaccard score = 0.79).

In other cases, text had been slightly modified to account for changes
in primary and secondary outcomes. Examples of these text changes were
found in topics 2 and 4; identified instances
related to sample size calculations for patient recruitment to different
studies.

Since studies in this dataset described planned analyses, we
hypothesised that some studies did not specify statistical methods because
they had yet to consult with a statistician. Targeted searches for ``statistician'' across all topics returned 1,277 matching studies,
with examples including ``A statistician employed by hospital was used''
and ``Pilot study at this point will use a statistician professionally
to determine sample size calculations as required''.

\caption{\label{tab:anzctr-example-boilerplate-sentence} Examples of boilerplate text from ANZCTR studies based on targeted keyword searches (sentence level). The number of matching papers was based on a Jaccard score of 0.75 or higher}
\centering
\begin{tabular}[t]{p{0.075\textwidth}p{0.15\textwidth}p{0.1\textwidth}p{0.85\textwidth}}
\hline
Topic & Search term &  Matching papers &  Example sentence from statistical methods section\\
\hline
\multicolumn{4}{l}{Statistical methods}\\
\hline
3 & ``t-test'' & 131 & Continuous Normally distributed variables will be compared using Student's t-test and reported as means standard deviation while non-normally distributed data will be compared using wilcoxon rank-sum tests and reported as medians inter-quartile-range\\
9 & ``model'' &  339 & Linear mixed models will be used to analyse the data.\\
10 & ``ANOVA'' & 86 & Post-hoc paired comparisons corrected for multiple comparisons will be performed if ANOVA reveal significant effects.\\
\hline
\multicolumn{4}{l}{Study designs}\\
4 & Patient & 375 & At a confidence level of 95\% and a precision around the estimate of 5\% a minimum of 73 patients must be included\\
5 & ``Pilot study'' & & \\
6 & ``Subject' & & \\
8 & ``'group'

\multicolumn{4}{l}{Qualitative analysis}\\
\hline
1 & & &\\
\hline
\multicolumn{4}{l}{Descriptive statistics}\\
\hline
6 & ``Mean $\pm$'' & 1,820 & Data are expressed as mean $\pm$ SEM.\\
\hline

3 & ``GraphPad'' &1,786  & All statistical analyses were performed using GraphPad Prism.\\
5 & ``SPSS'' & 1,104 & All statistical analyses were performed using SPSS 17.0 (SPSS Inc Chicago IL USA).\\
\hline

\multicolumn{4}{l}{Experiments}\\
\hline
10 & ``Experiment'' & 1,037 & All experiments were repeated at least three times.\\
\hline
\end{tabular}
\end{table}
\end{landscape}

\begin{table}
\caption{\label{tab:anzctr-example-boilerplate} Example boilerplate text from ANZCTR studies with the highest number of matches per topic (sentence level). The number of matching to each sentence was based on a Jaccard score of 0.75 or higher and a difference in word count of $\pm$ 3 words; $^{*}$ Most frequent match after excluding statements of statistical significance}
\centering
\begin{tabular}[t]{p{0.05\textwidth}p{0.8\textwidth}p{0.1\textwidth}}
\hline
Topic & Statistical methods text & Matching papers\\
\hline
1 & Significance will be set at p $<$ 0.05 & 13\\
 &  All analyses will be conducted on an intention-to-treat basis$^{*}$ & 8\\
\hline
2 & Analysis will be by intention-to-treat & 15\\
\hline
3 & A p-value $<$ 0.05 is considered statistically significant & 39\\
 & Continuous normally distributed variables will be compared using Student's t-test and reported as means \& standard deviation while non-normally distributed data will be compared using wilcoxon rank-sum tests and reported as medians inter-quartile-range$^{*}$ & 8\\
\hline
4 & Data will be analysed by two methods the Kaplan Meier method and per-protocol analysis & 27\\
\hline
5 & This is a pilot study only & 5\\
\hline
6 & Due to the exploratory nature of this first in human study no power or sample size calculations have been performed & 7\\
\hline
7 & Descriptive statistics & 7\\
\hline
8 & The level of statistical significance will be set at p $<$ 0.05 & 11\\
& Relationships between observations at different occasions will be modelled with an unstructured covariance matrix$^{*}$ & 9\\
\hline
9 & All analyses will be on an intention-to-treat basis & 24\\
\hline
10 & However allowing for a 50 percent attrition rate we will be required to recruit n=70 participants & 10\\

@@

>>>>

An ideal statistical analysis will use appropriate methods to draw
insights from the data and inform the research questions. Unfortunately
many current statistical analyses are far from ideal, with 
researchers often using the wrong methods, misinterpreting the results, or
failing to adequately check their assumptions \citep{2008, Leek2017}.
Some researchers take a ``mechanistic'' approach to statistics, copying
the few methods they know regardless of their appropriateness, and then
going through the motions of the analysis \citep{Stark2018}. Accepting this form of methodological illiteracy is at odds
with principles of scientific inquiry, yet continues to pervade published scientific research \citep{VanCalster2021}. 
This paradox has been exemplified during the COVID-19 pandemic, which has led to unprecendented levels of published research of largely 
poor quality due to biases in conduct and reporting \citep{Glasziou2020,Wynants2020}.

@@

We used two openly available data sources to find statistical methods
sections: study
protocols registered on the Australian and New Zealand Clinical Trials
Registry (ANZCTR) and research articles published in \emph{PLOS ONE}. Data sources were chosen as examples of common
research outputs that include descriptions of statistical methods that
were planned or used for analysing research outcomes.

@@

Data on all \emph{PLOS ONE} articles can be accessed via the PLOS
Application Programming Interface (API). This enabled us to conduct
searches of full-text articles and analyse data on articles' text
content and general attributes such as publication date and field(s) of
research. We applied a two-step
approach to identify statistical methods sections:

@@

Evidence of boilerplate text was assessed at the section and sentences levels using the Jaccard similarity index. 
We chose the Jaccard index as an easy to interpret measure, which summarises the similarity between two pieces of text by the
number of words common to each text (intersection), divided by the
number of words that appeared in either text (union). For both section-level and sentence-level analyses, text was tokenised at the word level. 
Instances of boilerplate text were defined by a Jaccard index of 0.8 or higher.

To reduce the number of pairwise comparisions, locality-sensitive hashing was implementing using the R package `textreuse' \citep{textreuseR}. For all topics, locality-sensitive hashing was applied using the same random seed, 
with 1,000 randomly selected minhashes broken into 200 equal-sized bands.

A similar approach was applied at the sentence level. In this case, we aimed to identify common sentences or phrases within statistical methods sections that were representative of each topic. 
Analyses considered keyword searches using dominant terms from word clouds and the reuse of sentences within the top 1,000 sections assigned to each topic, based on estimated coherence scores.

@@

We downloaded 28,008 studies. The numbers of excluded studies are shown
in Figure \ref{fig:consort-diagrams}. Of the 12,700 included studies,
9,523 (75\%) had a statistical methods section. The median length of
sections was 129 words with an inter-quartile range of 71 to 219 words. 
Sections for TODO studies were only one word, including ``ANOVA'',
``t-test'', ``SPSS'' and even ``SSPS''.

\begin{figure}[htbp]
\centering 
\includegraphics[width=0.49\linewidth]{figures/excluded_anzctr_missing}  \includegraphics[width=0.49\linewidth]{figures/excluded_plosone}
\caption{\label{fig:consort-diagrams}Search results for ANZCTR (left) and PLOS ONE (right).}
\end{figure}

@@

\begin{figure}
\centering 
\includegraphics[width=0.7\linewidth]{figures/anzctrwordclouds}
\caption{Word clouds for ten topics for statistical methods sections published in ANZCTR}\label{fig:anzctr-wordclouds}
\end{figure}

Figure \ref{fig:anzctr-wordclouds} shows word clouds for the ten topics identified by the NMF algorithm. Topics reflected sample size calculations (topics 2 and 5), 
study designs (topics 4, 5, 6 and 8),
quantitative methods (topics 3, 7, 9 and 10) and qualitative methods (topic 1). 
Common statistical methods for data analysis were t-testing, descriptive statistics, linear models and repeated measures ANOVA.

\begin{table}[htbp]
\centering
\caption{Results of boilerplate analysis applied to the ANZCTR dataset. The number of studies with Jaccard scores greater than or equal to 0.8 are presented. The number of exact matches is provided in brackets.}
\label{tab:anzctr-boilerplate-matches}
\begin{tabular}{lccc}
\hline
 &  & \multicolumn{2}{c}{Boilerplate text} \\
\cline{3-4}
Topic & Total studies & Section-level & {Sentence-level}\\
\hline
1 (Qualitative methods)& 854 & 53 (24) & 288 (206)\\
2 (Sample size calculations) & 1,834 & 68 (26) & 252 (172)\\
3 (Student's t-test) & 1,277 & 105 (45) & 552 (379)\\
4 (Efficacy and safety studies) & 909 & 85 (11) & 261 (203)\\
5 (Pilot studies) & 834 &  53 (34) & 148 (114)\\
6 (Safety and tolerability studies) & 524 & 51 (29) & 241 (186)\\
7 (Descriptive analysis) & 333 & 44 (39) & 69 (60)\\
8 (Intervention studies) & 1,020 & 33 (8) & 300 (203)\\
9 (Linear models) & 1,195 & 86 (33) & 436 (333)\\
10 (ANOVA) & 852 &  64 (29) & 270 (208)\\
\hline
\end{tabular}
\end{table}

Approximately 7$\%$ of all study sections were classified as boilerplate text (n = 642). One in ten sections associated with descriptive analysis were exact matches (Table \ref{tab:anzctr-boilerplate-matches}). Evidence of boilerplate text was stronger at the sentence level, with 29$\%$ of sections having at least one sentence flagged (n = 2,817).

Boilerplate text for selected topics emphasised definitions of statistical significance, combined with planned statistical methods. In topic 3 [TODO] Other topics simply stated the main analysis method, including ``descriptive statistics/descriptive analysis'' (topic 7, n = TODO), ``linear (mixed) models'' (topic 9, n = TODO) and ``repeated measures ANOVA'' (topic 10, n = TODO). Examples of common boilerplate sentences for describing statistical methods are in Table \ref{tab:anzctr-boilerplate-examples}. Full results for all topics are provided as Supplementary material.

In other cases, text had been slightly modified to account for changes in primary and secondary outcomes for a given study design. This behaviour was reflected in the high numbers of studies with boilerplate text at the section and sentence levels (Table \ref{tab:anzctr-boilerplate-matches}). Evidence of boilerplate text was strongest for topic 6, where matches aligned with the structure:

\begin{center}
here is some \underline{text} that i changed
\end{center}

Similar examples were identified for sample size calculations:

\begin{center}
here is some \underline{text} that i changed
\end{center}

Since studies in this dataset described planned analyses, we
hypothesised that some studies did not specify statistical methods because
they had yet to consult with a statistician. Targeted searches for ``statistician'' across all topics returned TODO matching studies,
with examples including ``A statistician employed by hospital was used''
and ``Pilot study at this point will use a statistician professionally
to determine sample size calculations as required''.

%\begin{landscape}
%\begin{table}
%
%\caption{\label{tab:anzctr-example-boilerplate-sentence} Examples of boilerplate text from ANZCTR studies based on targeted keyword searches (sentence level). The number of matching papers was based on a Jaccard score of 0.75 or higher}
%\centering
%\begin{tabular}[t]{p{0.075\textwidth}p{0.15\textwidth}p{0.1\textwidth}p{0.85\textwidth}}
%\hline
%Topic & Search term &  Matching papers &  Example sentence from statistical methods section\\
%\hline
%\multicolumn{4}{l}{Statistical methods}\\
%\hline
%3 & ``t-test'' & 131 & Continuous Normally distributed variables will be compared using Student's t-test and reported as means standard deviation while non-normally distributed data will be compared using wilcoxon rank-sum tests and reported as medians inter-quartile-range\\
%9 & ``model'' &  339 & Linear mixed models will be used to analyse the data.\\
%10 & ``ANOVA'' & 86 & Post-hoc paired comparisons corrected for multiple comparisons will be performed if ANOVA reveal significant effects.\\
%\hline
%\multicolumn{4}{l}{Study designs}\\
%4 & ``Patient'' & 375 & At a confidence level of 95\% and a precision around the estimate of 5\% a minimum of 73 patients must be included.\\
%5 & ``Pilot study'' & 16 & This is a pilot study.\\
%6 & ``Subject' & & \\
%8 & ``'group' & & \\
%\hline
%\multicolumn{4}{l}{Study designs}\\
%2 & ``'80 percent power'
%\hline
%\multicolumn{4}{l}{Qualitative analysis}\\
%\hline
%1 & & &\\
%\hline
%\multicolumn{4}{l}{Descriptive statistics}\\
%\hline
%7 & ``Descriptive'' & 33 & Analyses will be of a descriptive nature.\\
%\hline
%
%\multicolumn{4}{l}{Experiments}\\
%\hline
%10 & ``Experiment'' & 1,037 & All experiments were repeated at least three times.\\
%\hline
%\end{tabular}
%\end{table}
%\end{landscape}

\begin{table}
\caption{\label{tab:anzctr-example-boilerplate} Example boilerplate text from ANZCTR studies with the highest number of matches per topic (sentence level). The number of matching to each sentence was based on a Jaccard score of 0.75 or higher and a difference in word count of $\pm$ 3 words; $^{*}$ Most frequent match after excluding statements of statistical significance}
\centering
\begin{tabular}[t]{p{0.05\textwidth}p{0.8\textwidth}p{0.1\textwidth}}
\hline
Topic & Statistical methods text & Matching papers\\
\hline
1 & Significance will be set at p $<$ 0.05 & 13\\
 &  All analyses will be conducted on an intention-to-treat basis$^{*}$ & 8\\
\hline
2 & Analysis will be by intention-to-treat & 15\\
\hline
3 & A p-value $<$ 0.05 is considered statistically significant & 39\\
 & Continuous normally distributed variables will be compared using Student's t-test and reported as means \& standard deviation while non-normally distributed data will be compared using wilcoxon rank-sum tests and reported as medians inter-quartile-range$^{*}$ & 8\\
\hline
4 & Data will be analysed by two methods the Kaplan Meier method and per-protocol analysis & 27\\
\hline
5 & This is a pilot study only & 5\\
\hline
6 & Due to the exploratory nature of this first in human study no power or sample size calculations have been performed & 7\\
\hline
7 & Descriptive statistics & 15\\
\hline
8 & The level of statistical significance will be set at p $<$ 0.05 & 11\\
& Relationships between observations at different occasions will be modelled with an unstructured covariance matrix$^{*}$ & 9\\
\hline
9 & All analyses will be on an intention-to-treat basis & 24\\
\hline
10 & However allowing for a 50 percent attrition rate we will be required to recruit n=70 participants & 10\\

@@

\begin{table}[htbp]
\centering
\caption{Results of boilerplate analysis applied to the PLOS ONE dataset. The number of studies with Jaccard scores greater than or equal to 0.8 are presented. The number of exact matches is provided in brackets.}
\label{tab:plosr-boilerplate-matches}
\begin{tabular}{lccc}
\hline
 &  & \multicolumn{2}{c}{Boilerplate text} \\
\cline{3-4}
Topic & Total studies & Section-level & {Sentence-level}\\
\hline
1 (Student's t-test) & 3,784 & 344 () & )\\
2 () & 19,649 & 135 () &)\\
3 (GraphPad Prism) & 9,974 & 405 () &)\\
4 (ANOVA) & 10,212 & 452 () &)\\
5 (SPSS) & 9,648 & 287 () &)\\
6 (Descriptive analysis) & 4,764 & 331 () & \\
7 (Meta analysis) & 5,278&  65 () & \\
8 () & 42,261 &  116 () & \\
9 () & 6,195 & 260 () & \\
10 () & 4,693 & 336 () & \\
\hline
\end{tabular}
\end{table}
